---
title: "nycflights13"
author: "Jan Moskal"
date: "2023-08-19"
output: github_document
---

```{r setup, include=FALSE, message=FALSE, error=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Loading libraries and data

library(tidyverse)
library(nycflights13)
library(ggthemes)
library(gganimate)
library(scales)
library(patchwork)
library(extrafont)
library(plotly)

flights_df <- nycflights13::flights
weather_df <- nycflights13::weather
planes_df <- nycflights13::planes

flights_and_wheter_df <- left_join(flights, weather, by = c("origin", "year", "month", "day", "hour"))
flights_weather_planes <- left_join(flights_and_wheter_df, planes, by = "tailnum")
flights_weather_planes <- flights_weather_planes %>%
  select(-time_hour.y) %>%
  rename(built_year = year.y)
```


 What is the average departure and arrival delay?
```{r}
flights %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm = TRUE),
            avg_arr_delay = mean(arr_delay, na.rm = TRUE), 
            median_dep_delay = median(dep_delay, na.rm = TRUE),
            median_arr_delay = median(arr_delay, na.rm = TRUE))

```


 Is there any relationship between delay and time of day, day of the week, or month?
```{r}
# ------------------------------------------------------------------------------------------------

flights %>%
  group_by(hour = sched_dep_time %/% 100) %>%
  summarize(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = hour, y = avg_delay)) +
  geom_histogram(stat = "identity")
# Late night and early morning flights are less often delayed.

flights %>%
  group_by(month) %>%
  summarise(avg_month_delay = mean(arr_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = month <- factor(month, labels = month.name), y = avg_month_delay, fill = as.factor(month))) +
  geom_bar(stat = "identity") +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 15, color = "red", hjust = 0.5)
  ) +
  labs(
    title = "Average delay in each month",
    x = NULL,
    y = "Average delay"
  )
# Flights are delayed most in holidays and December and least in autumn
```


 Is there a correlation between route length and delay?
 How does seasonality impact flight delays?
```{r}
# ------------------------------------------------------------------------------------------------
flights %>%
  mutate(distance = cut(nycflights13::flights$distance,
    breaks = seq(0, 5000, by = 1000),
    labels = c("1000", "2000", "3000", "4000", "5000")
  )) %>%
  group_by(distance) %>%
  summarise(avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = distance, y = avg_arr_delay)) +
  geom_bar(stat = "identity")


flights %>%
  mutate(distance = cut(nycflights13::flights$distance,
    breaks = seq(0, 5000, by = 1000),
    labels = c("1000", "2000", "3000", "4000", "5000")
  )) %>%
  group_by(distance) %>%
  summarise(avg_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = distance, y = avg_dep_delay)) +
  geom_bar(stat = "identity")

# Long-distance flights often make up for the flight delay during the flight
# No significant relationship between dep_delay and distance
```


Which airlines have the highest number of flights?
```{r}
flights %>% 
  group_by(carrier) %>% 
  count(sort = TRUE) %>% 
  ggplot(aes(x = fct_reorder(carrier, n), y = n, fill = carrier)) +
  geom_bar(stat= "identity") +
  theme(legend.position = "none") +
  labs(x = NULL, 
       y = "Count") +
  scale_y_continuous(labels = label_number_si())
  
```


 Is there a relationship between the airline and delay?
```{r}
# ------------------------------------------------------------------------------------------------

flights %>%
  group_by(carrier) %>%
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(carrier, avg_delay), y = avg_delay, fill = carrier)) +
  geom_bar(stat = "identity")

# Flight delays depends on carriers
```

 Is there a relationship between the origin and delay?
```{r}
flights %>%
  group_by(origin) %>%
  summarize(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = origin, y = avg_delay, fill = origin)) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none")

# Flight delay kinda depends on origin
```


 What are the most popular flight routes in the dataset?
```{r}
flights %>% 
  group_by(origin, dest) %>% 
  count(sort = TRUE)

flights %>% 
  group_by(origin, dest) %>% 
  count(sort = TRUE) %>% 
  ungroup() %>% 
  mutate(x = row_number()) %>% 
  ggplot(aes(x = x, y = n)) +
  geom_histogram(stat = "identity")

```


 Which routes experience the most delays?
```{r}
# Total dep_delay  for each route
flights %>% 
  group_by(origin, dest) %>% 
  summarise(total_delay = sum(dep_delay, na.rm = TRUE)) %>% 
  arrange(desc(total_delay))

# Total arr_delay for each route
flights %>% 
  group_by(origin, dest) %>% 
  summarise(total_delay = sum(arr_delay, na.rm = TRUE)) %>% 
  arrange(desc(total_delay))

# Average dep_delay for each route
flights %>% 
  group_by(origin, dest) %>% 
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>% 
  arrange(desc(avg_delay))

# Average arr_delay for each route
flights %>% 
  group_by(origin, dest) %>% 
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  arrange(desc(avg_delay))
```


 Which airport has the most flights?
```{r}
flights %>% 
  count(origin, sort = TRUE)
```
 
 
 Are there airports where delays are more frequent?
```{r}
flights %>% 
  mutate(delayed = arr_delay > 0) %>% 
  group_by(origin) %>% 
  summarise(prop_delayed = mean(delayed, na.rm = TRUE)) %>% 
  ggplot(aes(x = origin, y = prop_delayed, fill = origin)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(name = "Proportion of delayed to all", labels = label_percent()) +
  theme(legend.position = "none") +
  labs(x = "Origin")

flights %>% 
  group_by(origin) %>% 
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE),
            total_delay = sum(arr_delay, na.rm = TRUE))

```
 
 
 Which airports are busiest at different times of the day?
```{r}
flights %>% 
  group_by(hour) %>% 
  ggplot(aes(x = hour)) +
  geom_bar(aes(fill = origin)) +
    facet_wrap(~ origin) +
  theme(legend.position = "none")
```


 What are the most commonly used aircraft types in the area?
```{r}
planes_df <- planes %>% 
  select(tailnum, type)
flights_and_planes <- left_join(flights, planes_df, by = "tailnum")

flights_and_planes %>% 
  count(type, sort = TRUE)
```
 
 Do different aircraft types have different delay rates?
```{r}
flights_and_planes %>% 
  drop_na(type) %>% 
  group_by(type) %>% 
  summarise(avg_arr_delay = mean(arr_delay, na.rm = TRUE),
            avg_dep_delay = mean(dep_delay, na.rm = TRUE))
```
 

Does the age of aircraft impact their performance?
```{r}
 planes_df1 <- planes %>% 
  select(tailnum, type, year) %>% 
  rename(plane_year = year)
flights_and_planes1 <- left_join(flights, planes_df1, by = "tailnum")


flights_and_planes1 %>% 
  mutate(speed = distance / (air_time/ 60)) %>% 
  group_by(plane_year) %>% 
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE),
            avg_speed = mean(speed, na.rm = TRUE)) %>% 
  arrange(desc(avg_delay))
```


 What were the weather conditions on different quarters, and what impact did they have on flight delays?
 Was there a correlation between delays and specific weather variables such as precipitation, wind, or visibility?
```{r}
weather %>% 
  group_by(quarter = case_when(
   month %in% c(12, 1, 2) ~ "Winter",
   month %in% c(9, 10, 11) ~ "Autumn",
   month %in% c(8, 7, 6) ~ "Summer",
   month %in% c(3, 4, 5) ~ "Spring",
  )) %>% 
  summarise(avg_visiblity = mean(visib, na.rm = TRUE)) %>% 
  arrange(desc(avg_visiblity))

weather %>% 
  group_by(quarter = case_when(
   month %in% c(12, 1, 2) ~ "Winter",
   month %in% c(9, 10, 11) ~ "Autumn",
   month %in% c(8, 7, 6) ~ "Summer",
   month %in% c(3, 4, 5) ~ "Spring",
  )) %>% 
  summarise(avg_precip = mean(precip, na.rm = TRUE),
            total_precip = sum(precip, na.rm = TRUE)) %>% 
  arrange(desc(avg_precip))

weather %>% 
  group_by(quarter = case_when(
   month %in% c(12, 1, 2) ~ "Winter",
   month %in% c(9, 10, 11) ~ "Autumn",
   month %in% c(8, 7, 6) ~ "Summer",
   month %in% c(3, 4, 5) ~ "Spring",
  )) %>% 
  summarise(avg_wind_speed = mean(wind_speed, na.rm = TRUE)) %>% 
  arrange(desc(avg_wind_speed))


# Visibilty is worst at Winter, rainfall are largest at Summer, windspeed is lowest at Summer

weather_df <- weather %>% 
  select(origin:hour, wind_speed, precip, visib)

flights_weather <- left_join(flights, weather_df, by = c("origin", "year", "month", "day", "hour"))

flights_weather %>% 
  ggplot(aes(x = precip, y = arr_delay)) +
  geom_point()

# I can not see relationship between precip and delay of flight

flights_weather %>% 
  drop_na(visib) %>% 
  group_by(visib) %>% 
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>% 
  ggplot(aes(x = as.factor(visib), y = avg_delay)) +
  geom_bar(stat = "identity")

flights_weather %>% 
  drop_na(precip) %>% 
  group_by(precip) %>% 
  summarise(avg_delay = median(dep_delay, na.rm = TRUE)) %>% 
  ggplot(aes(x = precip, y = avg_delay)) +
  geom_histogram(stat = "identity") 

flights_weather %>% 
  drop_na(wind_speed) %>% 
  group_by(wind_speed) %>% 
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>% 
  ggplot(aes(x = wind_speed, y = avg_delay)) +
  geom_histogram(stat = "identity") 
```
 